
\documentclass{article}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} %
\usepackage{multirow}
\usepackage{hyperref}

\usepackage{url}
\usepackage[section]{placeins}
\usepackage{algorithmic}
\usepackage{amsfonts}       %
\usepackage{nicefrac}       %
\usepackage{xcolor}         %
\usepackage{wrapfig}
\usepackage{listings}
\usepackage{pifont}
\usepackage{dsfont}


\newcommand{\theHalgorithm}{\arabic{algorithm}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}

\usepackage[accepted]{icml2023}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

\usepackage[capitalize,noabbrev]{cleveref}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newenvironment{proof-sketch}{\noindent{\textit{Sketch of proof.}}}


\usepackage[textsize=tiny]{todonotes}

\input{math_commands.tex}
\newcommand{\lc}[1]{{\color{red}{[\textbf{lc}: #1]}}}

\newcommand{\junz}[1]{{\color{red}{[\textbf{jz: }#1]}}}
\newcommand{\hangx}[1]{{\color{purple}{[[\textbf{hang: }#1]]}}}
\newcommand{\huayu}[1]{{\color{blue}{[[\textbf{chy: }#1]]}}}

\newcommand\Tstrut{\rule{0pt}{2.6ex}}         %
\newcommand\Bstrut{\rule[-0.9ex]{0pt}{0pt}}   %

\icmltitlerunning{Contrastive Energy Prediction for Exact Energy-Guided Diffusion Sampling in Offline Reinforcement Learning}

\begin{document}

\twocolumn[
\icmltitle{Contrastive Energy Prediction for Exact Energy-Guided \\ 
Diffusion Sampling in Offline Reinforcement Learning} 




\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}

\icmlauthor{Cheng Lu}{equal,thu}
\icmlauthor{Huayu Chen}{equal,thu}
\icmlauthor{Jianfei Chen}{thu}
\icmlauthor{Hang Su}{thu}
\icmlauthor{Chongxuan Li}{ruc}
\icmlauthor{Jun Zhu}{thu,pazhou}


\end{icmlauthorlist}

\icmlaffiliation{thu}{Dept. of Comp. Sci. \& Tech., Institute for AI, BNRist Center, Tsinghua-Bosch Joint ML Center, THBI Lab, Tsinghua University}
\icmlaffiliation{ruc}{Gaoling School of AI, Renmin University of China; Beijing Key Lab of Big Data Management and Analysis Methods, Beijing, China}
\icmlaffiliation{pazhou}{Pazhou Lab (Huangpu), Guangzhou, China}

\icmlcorrespondingauthor{Jun Zhu}{dcszj@tsinghua.edu.cn}

\icmlkeywords{Diffusion Models, Guided Sampling, Offline Reinforcement Learning}

\vskip 0.3in
]



\printAffiliationsAndNotice{\icmlEqualContribution} %

\begin{abstract}
Guided sampling is a vital approach for applying diffusion models in real-world tasks that embeds human-defined guidance during the sampling procedure. This paper considers a general setting where the guidance is defined by an (unnormalized) energy function. The main challenge for this setting is that the intermediate guidance during the diffusion sampling procedure, which is jointly defined by the sampling distribution and the energy function, is unknown and is hard to estimate. To address this challenge, we propose an exact formulation of the intermediate guidance as well as a novel training objective named contrastive energy prediction (CEP) to learn the exact guidance. Our method is guaranteed to converge to the exact guidance under unlimited model capacity and data samples, while previous methods can not. We demonstrate the effectiveness of our method by applying it to offline reinforcement learning (RL). Extensive experiments on D4RL benchmarks demonstrate that our method outperforms existing state-of-the-art algorithms. We also provide some examples of applying CEP for image synthesis to demonstrate the scalability of CEP on high-dimensional data. Code is available at \url{https://github.com/thu-ml/CEP-energy-guided-diffusion}.







\end{abstract}


\input{1-intro}
\input{2-background}
\input{3-method}
\input{4-comparison}
\input{5-offlineRL}
\input{6-image}


\section{Conclusion}
In this work, we formally consider the problem of sampling by diffusion models pretrained on a data distribution but the target sampling distribution is edited by an energy function. We show that this can be achieved by adding additional energy guidance to the original sampling procedure. We further propose a novel training objective named contrastive energy prediction for training an energy model to estimate such guidance. Our proposed CEP guidance is exact compared with previous energy guidance methods in the sense that it can guarantee convergence to the desired distribution. We apply our proposed method to several downstream tasks in order to demonstrate its effectiveness and scalability. Experimental results show that our method outperforms existing guidance methods in offline RL and is roughly on par with the classic classifier guidance in conditional image synthesis.


\section*{Acknowledgements}
This work was supported by the National Key Research and Development Program of China
(2020AAA0106302); 
NSF of China Projects (Nos. 62061136001, 61620106010, 62076145, U19B2034, U1811461, U19A2081, 6197222, 62106120, 62076145); a grant from Tsinghua Institute for Guo Qiang; the High Performance Computing Center, Tsinghua University. J.Z was also supported by the New Cornerstone Science Foundation through the XPLORER PRIZE. C. Li was also sponsored by the Beijing Nova Program.


\input{0-backup}






\nocite{langley00}

\bibliography{ref}
\bibliographystyle{icml2023}


\newpage
\appendix
\onecolumn


\input{0-appendix}

\end{document}
